---
title: "Merge MG22"
author: "Emily Kozik"
date: "1/6/2021"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective
This is a single cell RNA dataset of microglia in postmortem brain tissue. 
There are 3 samples from 2 donors. 
- Samples MG_22 were isolated as sorted microglia from Hippocampus (HIPP) and medial frontal gyrus (MFG)
- Sample 17_016 was isolated as single nuclei from unsorted medial frontal gyrus (MFG). 

We will merge the two fresh microglia samples and keep the nuclei as a separate dataset.

## Setup
```{r libraries}
#load libaries
library(dplyr)
library(Seurat)
library(patchwork)

print(paste("Seurat ", packageVersion("Seurat")))

```

Load the fresh microglia samples
```{r Loading Data}
load("~/Raj_Lab/Microglia/seurat_object_MG_22_HIPP.Rdata")
MG22HIPP <- seurat_object_MG_22_HIPP
rm(seurat_object_MG_22_HIPP)

load("~/Raj_Lab/Microglia/seurat_object_MG_22_MFG.Rdata")
MG22MFG <- seurat_object_MG_22_MFG
rm(seurat_object_MG_22_MFG)
```

## Look at each sample alone

This is the microglia from the hippocampus
```{r metadata MG22HIPP}
MG22HIPP
head(MG22HIPP@meta.data)
```

This is the microglia from the medial frontal gyrus
```{r metadata MG22MFG}
MG22MFG
head(MG22MFG@meta.data)
```

## Merge
```{r Merge MG22}
print("MG22HIPP")
MG22HIPP
print("MG22MFG")
MG22MFG
MG22 <- merge(MG22HIPP, y = MG22MFG, add.cell.ids = c("HIPP", "MFG"), project = "MG22_Microglia")
saveRDS(MG22, file = "MG22_combined.rds")
print("Combined: MG22")
MG22
```

```{r Features and Counts}
data.frame("median transcripts" = c("MG22HIPP" = median(MG22HIPP@meta.data$nCount_RNA),
                                    "MG22MFG" = median(MG22MFG@meta.data$nCount_RNA),
                                    "Combined" = median(MG22@meta.data$nCount_RNA)), 
            "median genes" = c("MG22HIPP" = median(MG22HIPP@meta.data$nFeature_RNA),
                               "MG22MFG" = median(MG22MFG@meta.data$nFeature_RNA),
                               "Combined" = median(MG22@meta.data$nFeature_RNA)))
```

